---
title: "Regression with a binary dependent variable"
author: "Zahid Asghar<br><br> School of Economics, QAU, Islamabad"
logo: "QAU-Logo.jpg"
toc: true
sidebar: false
page-layout: full
format:
  revealjs: 
    theme: [default, custom.scss]
    code-link: true
    code-fold: true   
editor: visual
---

## Regression with a binary dependent variable

So far the dependent variable ($Y$) has been continuous:

-   district-wide average test score
-   economic growth rate
-   earnings

What if $Y$ is binary?

-   $Y$ = get into college, or not; $X$ = years of education

\- $Y$ = person smokes, or not; $X$ = income

\- $Y$ = mortgage application is accepted, or not; $X$ = income, house characteristics, marital status, race

##  {.scrollable}

+----------------------------+--------------------------+-------------------------------+
| Topic                      | Dummy Dependent Variable | Description                   |
+============================+==========================+===============================+
| Labour Force Participation | Inlabourforce            | 0 if out of LF                |
|                            |                          |                               |
|                            |                          | 1 if in the LF                |
+----------------------------+--------------------------+-------------------------------+
| Choice of Occupation       | Managerial               | 0 if not managerial           |
|                            |                          |                               |
|                            |                          | 1 if managerial               |
+----------------------------+--------------------------+-------------------------------+
| Firm Location              | Shoppingmall             | 0 if not in the shopping mall |
|                            |                          |                               |
|                            |                          | 1 if in the shopping mall     |
+----------------------------+--------------------------+-------------------------------+
| Union Membership           | Union                    | 0 if not a union member       |
|                            |                          |                               |
|                            |                          | 1 if a union member           |
+----------------------------+--------------------------+-------------------------------+
| Retirement                 | Retired                  | 0 if not retired              |
|                            |                          |                               |
|                            |                          | 1 if retired                  |
+----------------------------+--------------------------+-------------------------------+
| Use of Seat Belts          | Seatbeltused             | 0 if does not use seat belt   |
|                            |                          |                               |
|                            |                          | 1 if uses seat belt           |
+----------------------------+--------------------------+-------------------------------+

: Examples of Binary Dependent Variables

## Main concepts

We review the following concepts:

-   the linear probability model
-   the Probit model
-   the Logit model
-   maximum likelihood estimation of nonlinear regression models

## Scatter plot of the Boston HMDA data

```{r}
library(haven)
hmda <- read_stata("hmda_sw.dta")
# create the variables of interest
hmda$deny = ifelse(hmda$s7==3, 1, 0)
hmda$pi_ratio = hmda$s46/100
hmda$black = ifelse(hmda$s13==3, 1, 0)
library(ggplot2)
figure11_1 = ggplot(data=hmda, aes(x=pi_ratio, y=deny))
figure11_1 = figure11_1 +
geom_point(alpha=0.2, size=1) +
geom_smooth(method="lm", se=F) +
ylim(-0.4,1.4) + scale_x_continuous(breaks = seq(0,3,0.5)) +
labs(x="P/I Ratio", y="Deny") +
annotate("text",x=2.5,y=0.9,label="Mortgage denied") +
annotate("text",x=2.5,y=-0.1,label="Mortgage approved") +
theme_bw()
figure11_1
```

## Trimming the data

```{r}
# sample of the HMDA data with 150 observations
library(dplyr)
hmda_s1 = dplyr::sample_n(filter(hmda,s13==3 & s7==3 & s46/100<1), 35)
hmda_s2 = dplyr::sample_n(filter(hmda,s13==3 & s46/100<1), 15)
hmda_s3 = dplyr::sample_n(filter(hmda,s13==5 & s46/100<1), 50)
hmda_s = rbind(hmda_s1, hmda_s2, hmda_s3)
# create the variables of interest
hmda_s$deny = ifelse(hmda_s$s7==3, 1, 0)
hmda_s$pi_ratio = hmda_s$s46/100
hmda_s$black = ifelse(hmda_s$s13==3, 1, 0)
figure11_1mod = ggplot(data=hmda_s, aes(x=pi_ratio, y=deny))
figure11_1mod = figure11_1mod +
geom_point(size=2, shape=1) +
geom_smooth(method="lm", se=F) +
ylim(-0.4,1.4) + scale_x_continuous(breaks = seq(0,1,0.1)) +
labs(x="P/I Ratio", y="Deny") +
annotate("text",x=0.6,y=1.1,label="Mortgage denied") +
annotate("text",x=0.6,y=-0.1,label="Mortgage approved") +
annotate("label",x=0.3,y=0.5,label="LPM") +
theme_bw()
figure11_1mod
```

## LPM

Equation of the LPM regression line: $\begin{align} deny = \beta_0 + \beta_1 \times P/I\ ratio + u. \tag{11.1} \end{align}$

$\begin{align} \widehat{deny} = -\underset{(0.032)}{0.080} + \underset{(0.098)}{0.604} P/I \ ratio. \tag{11.2} \end{align}$

```{r}
library(lmtest)
library(sandwich)
LPM = lm(deny ~ pi_ratio, data=hmda)
coeftest(LPM, vcov = sandwich)

```

## LPM inclduing race variable

$\begin{align} \widehat{deny} =& \, -\underset{(0.029)}{0.091} + \underset{(0.089)}{0.559} P/I \ ratio + \underset{(0.025)}{0.177} black. \tag{11.3} \end{align}$

```{r}
LPM2 = lm(deny ~ pi_ratio + black, data=hmda)
coeftest(LPM2, vcov=sandwich)

```

## Logit and Probit model

```{r}
library(ggplot2)
ggplot(data.frame(x = c(-5,5)), aes(x=x)) + 
  stat_function(fun = pnorm, aes(colour = "Probit")) + 
  stat_function(fun = plogis, aes(colour = "Logit")) + 
  theme_bw() + 
  scale_colour_manual(name = "Function G",values = c("red", "blue")) +
  scale_y_continuous(name = "Pr(y = 1 | x)")
```

## Probit Model, Predicted Probabilities and Estimated Effects

Assume that $Y$ is a binary variable. The model $$Y= \beta_0 + \beta_1 + X_1 + \beta_2 X_2 + \dots + \beta_k X_k + u$$ with $P(Y = 1 \vert X_1, X_2, \dots ,X_k)\\ = \Phi(\beta_0 + \beta_1 + X_1 + \beta_2 X_2 + \dots + \beta_k X_k)$ is the population Probit model with multiple regressors $X_1, X_2, \dots, X_k$ and $\Phi(\cdot)$ is the cumulative standard normal distribution function.

## Logit regression

The population Logit regression function is $$\begin{align*}
  P(Y=1\vert X_1, X_2, \dots, X_k) \\=& \, F(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_k X_k) \\
  =& \, \frac{1}{1+e^{-(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_k X_k)}}.
\end{align*}$$ The idea is similar to Probit regression except that a different CDF is used: $F(x) = \frac{1}{1+e^{-x}}$

is the CDF of a standard logistically distributed random variable.

## Probit regression

$\begin{align} E(Y\vert X) = P(Y=1\vert X) = \Phi(\beta_0 + \beta_1 X). \tag{11.4} \end{align}$ $\beta_0 + \beta_1 X$ in (11.4) plays a role of a quantile $z$.Remember that $\Phi(z) = P(Z \leq z) \ , \ Z \sim \mathcal{N}(0,1)$ such that the Probit coefficient $\beta_1$ in (11.4) is the change in $z$ associated with a one unit change in $X$. Although the effect on $z$ of a change in $X$ is linear, the link between $z$ and the dependent variable $Y$ is nonlinear since $\Phi$ is a nonlinear function of $X$. .

## Probit chart code

```{r}
figure11_3mod = ggplot(data=hmda_s, aes(x=pi_ratio, y=deny))
figure11_3probit = figure11_3mod +
geom_point(shape=1, size=2) +
geom_smooth(method="glm",

method.args=list(family=binomial(link="probit")),
se=F) +
ylim(-0.4,1.4) +
scale_x_continuous(breaks = seq(0,1,0.1)) +
labs(x="P/I Ratio", y="Deny") +
annotate("label",x=0.3,y=0.5,label="Probit Regression") +
theme_bw()

```

## Probit regression {.r-fit-text}

$\begin{align} \widehat{P(deny\vert P/I \ ratio, black)} = \Phi (-\underset{(0.18)}{2.26} + \underset{(0.50)}{2.74} P/I \ ratio + \underset{(0.08)}{0.71} black) \end{align}$

```{r}
probit = glm(
deny ~ black + pi_ratio,
data=hmda,
family=binomial(link="logit"))
coeftest(probit, vcov=sandwich)

```

## LPM , Logit, Probit {.scrollable}

```{r}
# rename and create some variables for regression
hmda$hse_inc = hmda$s45/100
hmda$loan_val = hmda$s6/hmda$s50
hmda$ccred = hmda$s43
hmda$mcred = hmda$s42
hmda$pubrec = ifelse(hmda$s44>0,1,0)
hmda$denpmi = ifelse(hmda$s53==1,1,0)
hmda$selfemp = ifelse(hmda$s27a==1,1,0)
hmda$married = ifelse(hmda$s23a=="M",1,0)
hmda$single = ifelse(hmda$married==0,1,0)
hmda$hischl = ifelse(hmda$school>=12,1,0)
hmda$probunmp = hmda$uria
hmda$condo = ifelse(hmda$s51 == 1,1,0)
```

```{r}
# LPM
model_1 = lm(
deny ~ black + pi_ratio + hse_inc + loan_val + ccred +
mcred + pubrec + + denpmi + selfemp,
data=hmda)
# Logit
model_2 = glm(
deny ~ black + pi_ratio + hse_inc + loan_val + ccred +
mcred + pubrec + + denpmi + selfemp,
data=hmda,
family=binomial(link="logit"))
# Probit
model_3 = glm(
deny ~ black + pi_ratio + hse_inc + loan_val + ccred +
mcred + pubrec + + denpmi + selfemp,
data=hmda,
family=binomial(link="probit"))

```

```{r}
models<-list(model_1,model_2,model_3)
library(huxtable)
library(modelsummary)
modelsummary(models,estimate = "{estimate}{stars}", output="huxtable")

```

## [mroz data from AER/Woolridge](https://scpoecon.github.io/ScPoEconometrics/binary.html#the-linear-probability-model)

```{r}
data(mroz, package = "wooldridge")
plot(factor(inlf) ~ age, data = mroz, 
     ylevels = 2:1,
     ylab = "in labor force?")

```

## 

```{r}
LPM = lm(inlf ~ nwifeinc + educ + exper 
         + I(exper^2) + age +I(age^2) + kidslt6, mroz)
summary(LPM)

```

## 

```{r}
pr = predict(LPM)
plot(pr[order(pr)],ylab = "p(inlf = 1)")
abline(a = 0, b = 0, col = "red")
abline(a = 1, b = 0, col = "red")
```

## 

```{r}
library(dplyr)
library(ggplot2)
mroz %<>% 
  # classify age into 3 and huswage into 2 classes
  mutate(age_fct = cut(age,breaks = 3,labels = FALSE),
         huswage_fct = cut(huswage, breaks = 2,labels = FALSE)) %>%
  mutate(classes = paste0("age_",age_fct,"_hus_",huswage_fct))

LPM_saturated = mroz %>%
  lm(inlf ~ classes, data = .)

mroz$pred <- predict(LPM_saturated)

ggplot(mroz[order(mroz$pred),], aes(x = 1:nrow(mroz),y = pred,color = classes)) + 
  geom_point() + 
  theme_bw() + 
  scale_y_continuous(limits = c(0,1), name = "p(inlf)") +
  ggtitle("LPM in a Saturated Model is Perfectly Fine")

```

## Interpretation of coefficients {.scrollable}

```{r}
probit <- glm(inlf ~ age, 
                    data = mroz, 
                    family = binomial(link = "probit"))

logit <- glm(inlf ~ age, 
                    data = mroz, 
                    family = binomial(link = "logit"))
modelsummary::modelsummary(list("probit" = probit,"logit" = logit))
```

##  {.scrollable}

[shinyapp](https://floswald.shinyapps.io/marginal_effects_of_logit_probit/)

```{r}
library(mfx)
f <- "inlf ~ age + kidslt6 + nwifeinc" # setup a formula
glms <- list()
glms$probit <- glm(formula = f, 
                    data = mroz, 
                    family = binomial(link = "probit"))
glms$logit <- glm(formula = f, 
                    data = mroz, 
                    family = binomial(link = "logit"))
# now the marginal effects versions
glms$probitMean <- mfx::probitmfx(formula = f, 
                    data = mroz, atmean = TRUE)
glms$probitAvg <- mfx::probitmfx(formula = f, 
                    data = mroz, atmean = FALSE)
glms$logitMean <- mfx::logitmfx(formula = f, 
                    data = mroz, atmean = TRUE)
glms$logitAvg <- mfx::logitmfx(formula = f, 
                    data = mroz, atmean = FALSE)
modelsummary::modelsummary(glms, 
                           stars = TRUE,
                           gof_omit = "AIC|BIC",
                           title = "Logit and Probit estimates and marginal effects evaluated at mean of x or as sample average of effects")
```

## This is why R

```{r}
#| echo: true
#| fig-width: 18
#| fig-align: center
library(leaflet)
leaflet() %>%
  addTiles() %>%
  addMarkers(lng = 73.136946, lat =33.748294 ,
             popup = "School of Economics, QAU, Islamabad")
```
