---
title: "Chapter 9 R_replication_codes Nonlinear Regression"

author: "Zahid Asghar"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Uploaded data and required R packages

Data sets can be downloaded from [Stock and Watson](https://wps.pearsoned.com/aw_stock_ie_3/178/45691/11696965.cw/index.html). We are working here with California Test School data used in Chapter 4-9.  For further details, one may visit [Nishant Yonzan page](https://sites.google.com/view/nishantyonzan/home?authuser=0)

```{r,  warning=FALSE,message=FALSE}
library(tidyverse) ## Upload set of packages including dplyr and ggplot2


library(haven) ## To read data from STATA/SPSS we use *haven* package
library(tidyverse)
caschool<- read_dta("caschool.dta")
maschool <- read_dta("data/mcas.dta")

```

```{r}
library(qwraps2) ## A collection of wrapper functions for reproducible reports
options(dplyr.width = Inf) # to force R to print all values
# CA school mean and sd
ca_mean <-
t(
summarise(caschool,
"Test Score" = mean(testscr),
"Student-Teacher Ratio" = mean(str),
"% English Learners" = mean(el_pct),
"% Receiving Lunch Subsidy" = mean(meal_pct),
"Average District Income ($)" = mean(avginc*1000),
count=n()
)
)
ca_sd <-
t(
summarise(caschool,
"Test Score" = sd(testscr),
"Student-Teacher Ratio" = sd(str),
"% English Learners" = sd(el_pct),
"% Receiving Lunch Subsidy" = sd(meal_pct),
"Average District Income ($)" = sd(avginc*1000),
count=n()
))
```


```{r}
# MA school mean and sd
ma_mean <-
t(
summarise(maschool,
"Test Score" = mean(totsc4),
"Student-Teacher Ratio" = mean(tchratio),
"% English Learners" = mean(pctel),
"% Receiving Lunch Subsidy" = mean(lnch_pct),
"Average District Income ($)" =
mean(percap*1000),
count=n()
)
)
ma_sd <-
t(
summarise(maschool,
"Test Score" = sd(totsc4),
"Student-Teacher Ratio" = sd(tchratio),
"% English Learners" = sd(pctel),
"% Receiving Lunch Subsidy" = sd(lnch_pct),
"Average District Income ($)" = sd(percap*1000),
count=n()
)
)
##Table using kable:
library(knitr)
library(kableExtra)
##
## Attaching package: 'kableExtra'
## The following object is masked from 'package:dplyr':
##
## group_rows
kable(
cbind(ca_mean,ca_sd,ma_mean,ma_sd),
align = "c",
digits = 1,
"html",
booktabs = TRUE
) %>%
column_spec(
2:5, width = "4em") %>%
kable_styling(
bootstrap_options = "striped", full_width = FALSE) %>%
add_header_above(
c(" ", "Average" = 1, "SD"=1, "Average" = 1, "SD" = 1)) %>%
add_header_above(
c(" " , "California" = 2, "Massachusetts" = 2)) %>%

row_spec(5,hline_after = TRUE)
```

## Multiple Regression

```{r}
# load packages
library(lmtest)
library(sandwich)
```

Create binary variable for HiEL and also squares and cubes of _str_ variable.

#### For CA
```{r}
caschool$hiel<-ifelse(caschool$el_pct>median(caschool$el_pct),1,0)
caschool$str_squared=caschool$str^2
caschool$str_cubed<-caschool$str^3
caschool$inc_squared<-caschool$avginc^2
caschool$inc_cubed<-caschool$avginc^3

```

#### For MA

```{r}
maschool$hiel = ifelse(maschool$pctel > median(maschool$pctel), 1, 0)
maschool$str_squared = maschool$tchratio^2
maschool$str_cubed = maschool$tchratio^3
maschool$inc_squared = maschool$percap^2
maschool$inc_cubed = maschool$percap^3
```

#### Changing names
```{r}
maschool$testscr = maschool$totsc4
maschool$str = maschool$tchratio
maschool$el_pct = maschool$pctel
maschool$meal_pct = maschool$lnch_pct
maschool$avginc = maschool$percap
```

## Regression

#### Regression models for CA

```{r}
CA_mod3<-lm(testscr~str+el_pct+meal_pct+avginc+inc_squared+inc_cubed, data=caschool)
CA_mod4<-lm(testscr~str+str_squared+str_cubed+el_pct+meal_pct+avginc+inc_squared+inc_cubed, data=caschool)
CA_mod5 <- lm(
testscr ~ str + meal_pct + hiel + hiel*str + avginc + inc_squared + inc_cubed,
data = caschool
)


```

#### Regression for MA

```{r}

MA_mod3 = lm(
testscr ~ str + el_pct + meal_pct + avginc + inc_squared + inc_cubed,
data = maschool
)
MA_mod4 = lm(
testscr ~ str + str_squared + str_cubed + el_pct + meal_pct + avginc + inc_squared + inc_cubed,
data = maschool
)
MA_mod5 = lm(
testscr ~ str + meal_pct + hiel + hiel*str + avginc + inc_squared + inc_cubed,
data = maschool
)
```

```{r}
library(huxtable)
huxreg("CA(1)"=CA_mod3, "CA(2)"=CA_mod4, "CA(3)"=CA_mod5,
"MA(1)"=MA_mod3, "MA(2)"=MA_mod4, "MA(3)"=MA_mod5,
coefs = c('STR'='str', 'STR^2'='str_squared', 'STR^3'='str_cubed' ,
'% English learners'='el_pct',
'% Eligible for free lunch'='meal_pct',
'HiEL (Binary, % English learners > median)'='hiel',
'HiEL X STR'='str:hiel','District income'='avginc',
'District income^2'='inc_squared', 'District income^3'='inc_cubed',
'Intercept'='(Intercept)'),
statistics = c('N' = 'nobs', 'R^2' = 'r.squared'),
stars = c(`*` = 0.1, `**` = 0.05)
)

```

```{r}
library(stargazer)
stargazer(CA_mod3, CA_mod4, CA_mod5,
MA_mod3, MA_mod4, MA_mod5,
dep.var.labels=c("California","Massachussets"),
type = 'text', align = TRUE,
          omit.stat=c("f", "ser"),
          column.sep.width = "-15pt")

```

