---
title: "Growth and Tradeshare"
author: "Zahid Asghar<br><br> School of Economics, QAU, Islamabad"
logo: "QAU-Logo.jpg"
format: 
  html:
    toc: true
    code-fold: true
editor: visual
---

## Growth and Trade Openess

Exclude `Malta` and carry out the following exercises. a. Run a regression of Growth on TradeShare, YearsSchool, Rev_Coups, Assassinations, and RGDP60. Construct a 95% confidence interval for the coefficient on TradeShare. Is the coefficient statistically significant at the 5% level? b. Test whether, taken as a group, YearsSchool, Rev_Coups, Assassinations, and RGDP60 can be omitted from the regression. What is the p-value of the F-statistic?

### 

Simple Linear Regression

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(haven)
library(kableExtra)
growth <- read_dta("Growth.dta")
library(modelsummary)
library(rstatix)
library(vtable)
vtable(growth)
```

Get an over of data as follows:

```{r}
glimpse(growth)
ggplot(growth)+aes(x=tradeshare,y=growth)+geom_point()
```

There seems an outlier and lets find out which value it is. Secondly what happens if we exclude this value as we have discussed OLS is sensitive to outliers.

```{r}
ggplot(growth)+aes(x=tradeshare,y=growth,color="red",label=country_name)+geom_point()+
geom_text(aes(label=ifelse(tradeshare>1.5,as.character(country_name),'')),hjust=0,vjust=0)+theme_minimal()
```

We observe that **Malta** is an outlier. Does this outlier affects **OLS** results it is very sensitive to outliers. We run both models : one with having Malta included and other by excluding Malta to analyze impact of tradeshare on growth rate.

```{r}
ggplot(growth)+aes(x=tradeshare,y=growth,color="red",label=country_name)+geom_point()+
geom_text(aes(label=ifelse(tradeshare>1.5,as.character(country_name),'')),hjust=0,vjust=0)+theme_minimal()+geom_smooth(method = "lm",se=FALSE)

```

```{r}
#| warning: false
#| message: false
library(huxtable)
Growth<-growth %>% filter(country_name!="Malta")
mod_malta<-lm(growth~tradeshare,data=growth)
mod_womalta<-lm(growth~tradeshare,data=Growth)
models<-list(mod_malta,mod_womalta)
tab<-modelsummary(models,estimate = "{estimate}{stars}", output="huxtable")
# customize table with `huxtable`
tab %>%
   
    # text color
    set_text_color(row = 4, col = 1:ncol(.), value = 'red')
```

#### Exclude `Malta` and carry out the following exercises.

a.  Run a regression of Growth on TradeShare, YearsSchool, Rev_Coups, Assassinations, and RGDP60. Construct a 95% confidence interval for the coefficient on TradeShare. Is the coefficient statistically significant at the 5% level?

```{r}
glimpse(Growth)
gr_model<- lm(growth~tradeshare+yearsschool+rev_coups+assasinations+rgdp60,data=Growth)
modelsummary(gr_model)
```

95% confidence interval can be calculated as follows:

**`r confint(gr_model, 'tradeshare', level=0.95)`**

#### b. Test whether, taken as a group, YearsSchool, Rev_Coups, Assassinations, and RGDP60 can be omitted from the regression. What is the p-value of the F-statistic?

```{r}
#| warning: false
#| message: false
library(car)
linearHypothesis(gr_model,c("yearsschool=0","rev_coups=0","assasinations=0","rgdp60=0"))
```
