---
title: " Regression Discontinuity Designs"
author: "Zahid Asghar"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: revealjs
institute: Quaid-i-Azam University, Islamabad
---

```{r setup, include=FALSE}

```

## Outline

-   Introduction to Regression Discontinuity Designs

-   Sharp and Fuzzy Regression Discontinuity Designs

-   Simulated Sharp and Fuzzy Regression Discontinuity Designs using R

-   Encourage the use of this technique and R

## Regression Discontinuity Designs (RDD)

When randomization is not feasible, how can we exploit implementation features of the program to measure its impact? - Answer: Quasi-experiments - Example: Regression Discontinuity Design.

*Regression discontinuity design* (RDD) is an important method that can be used for programs that have a continuous eligibility index with a clearly defined eligibility threshold (cuttoff score) to determine who is eligible and who is not.

-   The regression discontinuity design is a statistical approach that utilizes threshold based decision making to estimate causal estimates of different interventions.

-   Regression discontinuity is relatively simple to implement, transparent, and provides "real world" effects of treatments and policies.

-   Graphical presentation and statistical analysis of data.

-   Developed by Psychologists but later on in much use with labor economists

-   Despite frequent use threshold based decision making, regression discontinuity is underutilized.

## Types of RDD

-   With regard to Treatment Assignment:

    -   **Sharp Design**: Where the cut-point perfectly predicts who does/doesn't receive intervention

        -   Characteristics cant be perfectly manipulated by invidual

    -   **Fuzzy Design**: Where there exists "cross-overs" (for various reasons)

-   With regard to Analytic Approach:

    -   Non-parametric: local randomization approach
    -   Parametric: uses every observation in the sample

------------------------------------------------------------------------

-   Ideal for situations when there is a **cut-score** in which subjects above/below a threshold receive program resources or an intervention

-   The variable used to determine the threshold is called the **"rating" or "forcing"** variable. Examples: GPA, household income

-   Quasi-experimental design used to estimate the rigor and validity of a Randomized Control Trial (RCT)

-   In many fields, regression discontinuity are considered to be one of the most rigorous quasi-experimental methods

## Internal Validity

-   Imprecise Control

    -   No one single entity or factor should determine a rating score. For a RDD to yield an unbiased LATE estimate at the cuttoff, it is important that the eligibility index not be manipulated around the cutoff so that an individual can change treatment control status.

-   Clear discontinuity for treatment status at the cut-point

## The Scenario

-   Ehsaas scholarship criteria for grant of support is Rs.45,000 income of parents *prior to viewing any actual data*. A threshold of Rs.45,000/ month income is established: students below this cut-point will receive scholarshio, those above will not.
-   Besides assistance, one may want to determine if scholarship proves effective in relation to final grade.

## Simulation Data

-   Institutional Data:

    -   Taken from supplemental instruction STEM report, N=589
    -   Scores are adjusted on outcome for demonstration purposes

## Model

$$ Y_i=\alpha+\beta_0 T_i+\beta_1 r_i+\epsilon$$ $\alpha$ is Estimated marginal mean ,$\beta_0$ is marginal impact of program, $T_i$ is Treatment status (0,1), $|beta_1$ is Relationship between scholarship and Final Grade, $r_i$ is Individual scores on the rating variable (Scholarship amount; centered)

## Sharp Designs

```{r, warning=FALSE, message=FALSE}

# generate some sample data
W <- runif(1000, -1, 1)
y <- 3 + 2 * W + 10 * (W>=0) + rnorm(1000)
# load the package 'rddtools'
library(rddtools)

# construct rdd_data 
data <- rdd_data(y, W, cutpoint = 0)
library(ggplot2)
ggplot(data)+aes(x=W,y=y)+geom_point()
# estimate the sharp RDD model
rdd_mod <- rdd_reg_lm(rdd_object = data, 
                      slope = "same")
summary(rdd_mod)
```

```{r }
# plot the sample data

plot(data,
     col = "steelblue",
     cex = 0.35, 
     xlab = "W", 
     ylab = "Y")

# plot the RDD model along with binned observations
plot(rdd_mod,
     cex = 0.35, 
     col = "steelblue", 
     xlab = "W", 
     ylab = "Y")
```

## Fuzzy Designs

```{r plot}
library(MASS)
# generate sample data
mu <- c(0, 0)
sigma <- matrix(c(1, 0.7, 0.7, 1), ncol = 2)

set.seed(1234)
d <- as.data.frame(mvrnorm(2000, mu, sigma))
colnames(d) <- c("W", "Y")

# introduce fuzziness
d$treatProb <- ifelse(d$W < 0, 0, 0.8)

fuzz <- sapply(X = d$treatProb, FUN = function(x) rbinom(1, 1, prob = x))

# treatment effect
d$Y <- d$Y + fuzz * 2
# generate a colored plot of treatment and control group
plot(d$W, d$Y,
     col = c("steelblue", "darkred")[factor(fuzz)], 
     pch= 20, 
     cex = 0.5,
     xlim = c(-3, 3),
     ylim = c(-3.5, 5),
     xlab = "W",
     ylab = "Y")
# add a dashed vertical line at cutoff
abline(v = 0, lty = 2)
```

For more on these codes you can read [Ch.13 Introduction to Econometrics with R](https://www.econometrics-with-r.org/13-4-qe.html)
