---
format: 
  revealjs:
    self-contained: false
    slide-number: c/t
    width: 1600
    height: 900
    logo: "https://www.rstudio.com/wp-content/uploads/2018/10/RStudio-Logo-Flat.png"
    footer: "[zahid asghar](https://zahidasghar.com/)"
    theme: ["simple", "custom.scss"]
    echo: true
    multiplex: true
    code-fold: true
    chalkboard: true
    execute:
     freeze: auto  
    title-slide-attributes:
      data-background-color: "#447099"
editor: source
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}

library(tidyverse)

library(httr)
clrs <- MetBrewer::met.brewer(name = "Java")
clrs_lt <- colorspace::lighten(clrs, 0.9)
knitr::opts_chunk$set(fig.retina = 3, collapse = TRUE)
options(digits = 3, width = 75)
```

# Internal and external validity of <br> regression studies {background-color="`r clrs[3]`"}

## Goal of this study

-   [Internal versus External Validity]{.yellow}

. . .

-   Testing for Heteroskedasticity

. . .

> [A statistical analysis has internal validity if the statistical inference made about causal effects are valid for the considered population.]{.yellow}

. . .

> An analysis is said to have external validity if inferences and conclusion are valid for the studies' population and can be generalized to other populations and settings.

## Internal and External Validity of Regression

-   Internal validity is satisfied if the statistical inference about the causal effects are valid for the population being studied.
-   [External validity would need the inferences and conclusions to be generalized from the population and setting studied to other populations and setting]{.yellow}
-   Example: Think about the california test example.
-   [Is the slope, $\beta$**str** , unbiased and consistent?]{.yellow}
-   Is this a valid estimate for NY?, WV?, NM?,. . .

## Threats to internal validity

1.  **Omitted Variable Bias**

2.  **Misspecification of Functional Form of Regression Model**

3.  **Measurement Error**

4.  **Missing Data and Sample Selection**

5.  **Simultaneous Causality In each case, OLS assumption \[**$#1$**\]{yellow} is violated:** $E(u_i| X_{1i}, X_{2i}, \dots,X_{ki})\neq 0.$

## Omitted variable bias

-   **If you have the data for omitted variables:**
    -   [Be specific about your coefficient of interest.]{.yellow}
    -   Use a priori reason for adding variables.
    -   [Use statistical tests for questionable variables (t and F).]{.yellow}
    -   Provide all the potential specifications in tabular form.
-   **What if you don't have the data:**
    -   Panel data (next chapter)
    -   [IV method (chapter 12)]{.yellow}
    -   Randomized control experiments

## Misspecification of Functional Form of Regression Model

-   For continuous dependent variable: Use methods discussed in chapter 8 to modify functional forms.
-   For discrete or binary dependent variable: Chapter 11 (we will get there. . . ).

## 

Lets our model is $Y_i = X_i^2$ but one uses $Y_i=\beta_0+\beta_1X_{1i}+u_i$

::: panel-tabset
## Code

```{r}

# set seed for reproducibility
set.seed(5)
library(tidyverse)
# simulate data set
X <- runif(100, -5, 5)
Y <- X^2 + rnorm(100)
df<-cbind(X,Y)
df<-as.data.frame(df)
```

## Output

```{r}
#| eval: false
# estimate the regression function
ms_mod <- lm(Y ~ X)
ms_mod

```
:::

```{r}
# set seed for reproducibility
set.seed(5)
library(tidyverse)
# simulate data set
X <- runif(100, -5, 5)
Y <- X^2 + rnorm(100)
df<-cbind(X,Y)
df<-as.data.frame(df)
# estimate the regression function
ms_mod <- lm(Y ~ X)
ms_mod

```

## Plot this linear fit

```{r}
ggplot(df)+aes(X,Y)+geom_point()+geom_smooth(method = "lm",se=FALSE)

```

## Measurement Error in $X$

General regression model: $Y_i = \beta_0 + \beta_1X_{i} + u_{i}$ What you would like to measure is $X_i$, but what you do actually measure is $\overset{\sim}{X}_i$. Then the error is $X_i - \overset{\sim}{X}_i$

-   Regression model with mesurement error $\overset{\sim}{X}_i$ instead of $X_i$ $\begin{align*} Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + \underbrace{\beta_1 (X_i - \overset{\sim}{X}_i) + u_i}_{=v_i} \\ Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + v_i \end{align*}$

Rewrite $ν_i = \beta_1 (X_i - \overset{\sim}{X}_i)+u_i$

## 

$$\begin{align*}
  Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + \underbrace{\beta_1 (X_i - \overset{\sim}{X}_i) + u_i}_{=v_i} \\
  Y_i =& \, \beta_0 + \beta_1 \overset{\sim}{X}_i + v_i
\end{align*}$$

where here $\overset{\sim}{X}_i$ and $v_i$ are correlated resulting to slope coefficient $\hat{\beta_1}$ to be biased.

## 

The classical measurement error model assumes that the measurement error, $w_i$, has zero mean and that it is uncorrelated with the variable, $\overset{\sim}{X}_i$, and the error term of the population regression model, $u_i$: $\begin{equation} \overset{\sim}{X}_i = X_i + w_i, \ \ \rho_{w_i,u_i}=0, \ \ \rho_{w_i,X_i}=0 \end{equation}$ This holds $\begin{equation} \widehat{\beta}_1 \xrightarrow{p}{\frac{\sigma_{X}^2}{\sigma_{X}^2 + \sigma_{w}^2}} \beta_1 \tag{9.1} \end{equation}$

## 

$\sigma_{X}^2, \sigma_{w}^2 > 0$ so $\hat{\beta_1}$ smaller than 1.

1.  If there is no measurement error, $\sigma_{w}^2=0$ such that $\widehat{\beta}_1 \xrightarrow{p}{\beta_1}$ .

2.  If $\sigma_{w}^2 \gg \sigma_{X}^2$ we have $\widehat{\beta}_1 \xrightarrow{p}{0}$

This is the case if the measurement error is so large that there essentially is no information on $X$ in the data that can be used to estimate $\beta$.

## 

$\begin{align} (X, Y) \sim \mathcal{N}\left[\begin{pmatrix}50\\ 100\end{pmatrix},\begin{pmatrix}10 & 5 \\ 5 & 10 \end{pmatrix}\right] \tag{9.3} \end{align}$

$\begin{align*} Y_i =& \, 100 + 0.5 (X_i - 50) \\ =& \, 75 + 0.5 X_i. \tag{9.4} \end{align*}$

$\overset{\sim}{X_i} = X_i + w_i$ $w_i \overset{i.i.d.}{\sim} \mathcal{N}(0,10)$ $w_i$ is independent of $x_i$.

## 

```{r}
# set seed
set.seed(1)

# load the package 'mvtnorm' and simulate bivariate normal data
library(mvtnorm)
dat <- data.frame(
  rmvnorm(1000, c(50, 100), 
          sigma = cbind(c(10, 5), c(5, 10))))

# set columns names
colnames(dat) <- c("X", "Y")

```

##  Plots

```{r}
# estimate the model (without measurement error)
noerror_mod <- lm(Y ~ X, data = dat)

# estimate the model (with measurement error in X)
dat$X <- dat$X + rnorm(n = 1000, sd = sqrt(10))
error_mod <- lm(Y ~ X, data = dat)

# print estimated coefficients to console
noerror_mod$coefficients
error_mod$coefficients
```



```{r}
#| echo: false
# plot sample data
plot(dat$X, dat$Y, 
     pch = 20, 
     col = "steelblue",
     xlab = "X",
     ylab = "Y")

# add population regression function
abline(coef = c(75, 0.5), 
       col = "darkgreen",
       lwd  = 1.5)

# add estimated regression functions
abline(noerror_mod, 
       col = "purple",
       lwd  = 1.5)

abline(error_mod, 
       col = "darkred",
       lwd  = 1.5)
# add legend
legend("topleft",
       bg = "transparent",
       cex = 0.8,
       lty = 1,
       col = c("darkgreen", "purple", "darkred"), 
       legend = c("Population", "No Errors", "Errors"))

```

## plot in ggplot2

```{r}


# estimate the model (with measurement error in X)
dat$X1 <- dat$X + rnorm(n = 1000, sd = sqrt(10))
error_mod <- lm(Y ~ X, data = dat)
ggplot(dat)+aes(X,Y)+geom_smooth(aes(X, Y), data = dat, 
              method = "lm", se = FALSE, color = "red") +
  geom_smooth(aes(X1, Y), data = dat, 
              method = "lm", se = FALSE, color = "blue") +
  geom_point()

```

## Missing Data and Sample Selection

1.  **Missing at random: There is no bias, but it reduces our sample size.**
2.  [**Missing regressor values: Same as above.**]{.yellow}
3.  **Missing $Y$ due to selection process (sample selection bias): For example of this type of bias, think about why older, say, baseball players are on average better than younger ones.**

# 4. Sample Selection Bias

## 5. Simultaneous Causality

Simultaneous Causality Bias So far we have assumed that the changes in the independent variable $X$ are responsible for changes in the dependent variable $Y$ . When the reverse is also true, we say that there is simultaneous causality between $X$ and $Y$ . This reverse causality leads to correlation between $X$ and the error in the population regression of interest such that the coefficient on $X$ is estimated with bias.

# Summary {background-color="`r clrs[2]`"}

## Threats to Internal Validity of a Regression Study

The five primary threats to internal validity of a multiple regression study are:

-   Omitted variables

-   Misspecification of functional form

-   Errors in variables (measurement errors in the regressors)

-   Sample selection

-   Simultaneous causality

All these threats lead to failure of the first least squares assumption $E(u_i\vert X_{1i},\dots ,X_{ki}) \neq 0$

so that the OLS estimator is biased and inconsistent.

Furthermore, if one does not adjust for heteroskedasticity and/or serial correlation, incorrect standard errors may be a threat to internal validity of the study.

## Uploaded data and required R packages

Data sets can be downloaded from [Stock and Watson](https://wps.pearsoned.com/aw_stock_ie_3/178/45691/11696965.cw/index.html). We are working here with California Test School data used in Chapter 4-9.  For further details, one may visit [Nishant Yonzan page](https://sites.google.com/view/nishantyonzan/home?authuser=0)

```{r,  warning=FALSE,message=FALSE}
library(haven) ## To read data from STATA/SPSS we use *haven* package

caschool<- read_dta("data/caschool.dta")
maschool <- read_dta("data/mcas.dta")

```

```{r}
library(qwraps2) ## A collection of wrapper functions for reproducible reports
options(dplyr.width = Inf) # to force R to print all values
# CA school mean and sd
ca_mean <-
t(
summarise(caschool,
"Test Score" = mean(testscr),
"Student-Teacher Ratio" = mean(str),
"% English Learners" = mean(el_pct),
"% Receiving Lunch Subsidy" = mean(meal_pct),
"Average District Income ($)" = mean(avginc*1000),
count=n()
)
)
ca_sd <-
t(
summarise(caschool,
"Test Score" = sd(testscr),
"Student-Teacher Ratio" = sd(str),
"% English Learners" = sd(el_pct),
"% Receiving Lunch Subsidy" = sd(meal_pct),
"Average District Income ($)" = sd(avginc*1000),
count=n()
))
```


```{r}
# MA school mean and sd
ma_mean <-
t(
summarise(maschool,
"Test Score" = mean(totsc4),
"Student-Teacher Ratio" = mean(tchratio),
"% English Learners" = mean(pctel),
"% Receiving Lunch Subsidy" = mean(lnch_pct),
"Average District Income ($)" =
mean(percap*1000),
count=n()
)
)
ma_sd <-
t(
summarise(maschool,
"Test Score" = sd(totsc4),
"Student-Teacher Ratio" = sd(tchratio),
"% English Learners" = sd(pctel),
"% Receiving Lunch Subsidy" = sd(lnch_pct),
"Average District Income ($)" = sd(percap*1000),
count=n()
)
)
##Table using kable:
library(knitr)
library(kableExtra)
##
## Attaching package: 'kableExtra'
## The following object is masked from 'package:dplyr':
##
## group_rows
kable(
cbind(ca_mean,ca_sd,ma_mean,ma_sd),
align = "c",
digits = 1,
"html",
booktabs = TRUE
) %>%
column_spec(
2:5, width = "4em") %>%
kable_styling(
bootstrap_options = "striped", full_width = FALSE) %>%
add_header_above(
c(" ", "Average" = 1, "SD"=1, "Average" = 1, "SD" = 1)) %>%
add_header_above(
c(" " , "California" = 2, "Massachusetts" = 2)) %>%

row_spec(5,hline_after = TRUE)
```

## Multiple Regression

```{r}
# load packages
library(lmtest)
library(sandwich)
```

Create binary variable for HiEL and also squares and cubes of _str_ variable.

#### For CA
```{r}
caschool$hiel<-ifelse(caschool$el_pct>median(caschool$el_pct),1,0)
caschool$str_squared=caschool$str^2
caschool$str_cubed<-caschool$str^3
caschool$inc_squared<-caschool$avginc^2
caschool$inc_cubed<-caschool$avginc^3

```

#### For MA

```{r}
maschool$hiel = ifelse(maschool$pctel > median(maschool$pctel), 1, 0)
maschool$str_squared = maschool$tchratio^2
maschool$str_cubed = maschool$tchratio^3
maschool$inc_squared = maschool$percap^2
maschool$inc_cubed = maschool$percap^3
```

#### Changing names
```{r}
maschool$testscr = maschool$totsc4
maschool$str = maschool$tchratio
maschool$el_pct = maschool$pctel
maschool$meal_pct = maschool$lnch_pct
maschool$avginc = maschool$percap
```

## Regression

#### Regression models for CA

```{r}
CA_mod3<-lm(testscr~str+el_pct+meal_pct+avginc+inc_squared+inc_cubed, data=caschool)
CA_mod4<-lm(testscr~str+str_squared+str_cubed+el_pct+meal_pct+avginc+inc_squared+inc_cubed, data=caschool)
CA_mod5 <- lm(
testscr ~ str + meal_pct + hiel + hiel*str + avginc + inc_squared + inc_cubed,
data = caschool
)


```

#### Regression for MA

```{r}

MA_mod3 = lm(
testscr ~ str + el_pct + meal_pct + avginc + inc_squared + inc_cubed,
data = maschool
)
MA_mod4 = lm(
testscr ~ str + str_squared + str_cubed + el_pct + meal_pct + avginc + inc_squared + inc_cubed,
data = maschool
)
MA_mod5 = lm(
testscr ~ str + meal_pct + hiel + hiel*str + avginc + inc_squared + inc_cubed,
data = maschool
)
```

## Comparison{.scroll}

```{r}

library(huxtable)
CA_MA<-huxreg("CA(1)"=CA_mod3, "CA(2)"=CA_mod4, "CA(3)"=CA_mod5,
"MA(1)"=MA_mod3, "MA(2)"=MA_mod4, "MA(3)"=MA_mod5,
coefs = c('STR'='str', 'STR^2'='str_squared', 'STR^3'='str_cubed' ,
'% English learners'='el_pct',
'% Eligible for free lunch'='meal_pct',
'HiEL (Binary, % English learners > median)'='hiel',
'HiEL X STR'='str:hiel','District income'='avginc',
'District income^2'='inc_squared', 'District income^3'='inc_cubed',
'Intercept'='(Intercept)'),
statistics = c('N' = 'nobs', 'R^2' = 'r.squared'),
stars = c(`*` = 0.1, `**` = 0.05)
)

```

## {.scrollable}
:::{.r-fit-text}
```{r}
#| code-fold: false 
CA_MA
```
:::

